Why do species get a thin slice of $\pi$? Revisiting Lewontin's Paradox of Variation

### 2021-05-25 manuscript/manuscript.tex
updates to ms, code

@@

\begin{abstract} 
  Under neutral theory, genetic diversity in an equilibrium population
  is expected to increase with population size. However, observed levels of
  diversity across metazoans vary only two orders of magnitude, while census
  population sizes ($N_c$) are expected to vary over several. This unexpectedly
  narrow range of diversity is a longstanding enigma in evolutionary genetics
  known as Lewontin's Paradox of Variation (1974). Since Lewontin's
  observation, it has been argued that selection constrains diversity across
  species, yet tests of this hypothesis seem to fall short of explaining the
  orders-of-magnitude reduction in diversity observed in nature. In this work,
  I revisit Lewontin's Paradox and assess whether current models of linked
  selection are likely to constrain diversity to this extent.  To quantify the
  discrepancy between pairwise diversity and census population sizes across
  species, I combine genetic data from 172 metazoan taxa with estimates of
  census sizes from geographic occurrence data and population densities
  estimated from body mass.  Next, I fit the relationship between
  previously-published estimates of genomic diversity and these approximate
  census sizes to quantify Lewontin's Paradox. While previous across-taxa
  population genetic studies have avoided accounting for phylogenetic
  non-independence, I use phylogenetic comparative methods to investigate the
  diversity census size relationship, estimate phylogenetic signal, and explore
  how diversity changes along the phylogeny. I consider whether the reduction
  in diversity predicted by models of recurrent hitchhiking and background
  selection could explain the observed pattern of diversity across species.
  Since the impact of linked selection is mediated by recombination map length,
  I also investigate how map lengths vary with census sizes. I find species
  with large census sizes have shorter map lengths, leading these species to
  experience greater reductions in diversity due to linked selection. Even
  after using high estimates of the strength of sweeps and background
  selection, I find linked selection likely cannot explain the shortfall
  between predicted and observed diversity levels across metazoan species.
  Furthermore, the predicted diversity under linked selection does not fit the
  observed diversity--census-size relationship, implying that processes other
  than background selection and recurrent hitchhiking must be limiting
  diversity.

@@

A longstanding mystery in evolutionary genetics is that the observed levels of
genetic variation across sexual species are confined to an unexpectedly narrow
range. Under neutral theory, the average number of nucleotide differences
between lineages (pairwise diversity, $\pi$) is determined by the balance of
new mutations and their loss by genetic drift
\parencite{Kimura1964-ul,Malecot1948-zv,Wright1931-fl}. In particular, the
expected diversity at neutral sites in a panmictic population of $N_c$ diploids
is expected to be $\pi \approx 4N_c \mu$, where $\mu$ is per generation
mutation rate. Given that metazoan germline mutation rates only differ 10-fold
($10^{-8}$--$10^{-9}$, \cite{Kondrashov2010-fi,Lynch2010-ki}), and census sizes
vary over several orders of magnitude, one would expected under neutral theory
that heterozygosity should also vary over several orders of magnitude. However,
early allozyme surveys revealed that heterozygosity levels across a wide range
of species varied just an order of magnitude \parencite[][p.
208]{Lewontin1974-jb}; this anomaly is known as Lewontin's ``Paradox of
Variation". With modern sequencing-based estimates of $\pi$ across taxa ranging
over only three orders of magnitude ($0.01$--$10\%$, \cite{Leffler2012-zj}),
Lewontin's paradox has persisted unresolved through the genomics era.

From the beginning, explanations for Lewontin's Paradox have been framed in
terms of the neutralist--selectionist controversy
\parencite{Lewontin1974-jb,Kimura1984-ia,Gillespie1991-qa,Gillespie2001-mv}.
The neutralist view is that beneficial alleles are sufficiently rare and
deleterious alleles removed sufficiently quickly, that levels of genetic
diversity are shaped predominantly by genetic drift and mutation
\parencite{Kimura1984-ia}. Specifically, \emph{non-selective} processes
decouple the effective population size implied by observed levels of diversity
$\widehat{\pi}$, $\widetilde{N}_e = \nicefrac{\widehat{\pi}}{4\mu}$, from the
census size, $N_c$. By contrast, the selectionist view is that the direct and
indirect effects of linked selection suppress diversity levels across taxa,
specifically because the impact of linked selection is greater in large
populations. Undoubtedly, these opposing views represent a false dichotomy, as
population genomic studies have uncovered complex demographic histories that
impact diversity within a species (e.g. \cite{Zhao2013-vd,Palkopoulou2015-bg}),
as well as evidence that selection depresses genome-wide diversity (e.g.
\cite{Elyashiv2016-vt,Begun1992-ey,Aguade1989-jx,McVicker2009-ax}).

@@

While chance demographic events like bottlenecks and recent expansions have
long-term impacts on diversity (equilibrium is reached on the order of size of
the population), characteristics of the breeding structure such as high
variance ($V_w$) or skew in reproductive success  also suppress diversity below
the levels predicted by the census size \parencite{Wright1938-tv}. In species
like marine animals, females are highly fecund, and dispersing larvae face
extremely low survivorship, leading to high variance in reproductive success
\parencite{Waples2018-kb,Waples2013-wi,Hedgecock2011-ku,Hauser2008-fd}. Such
``sweepstakes" reproductive systems can lead to remarkably small ratios of
effective population size to census population size (e.g. $\nicefrac{N_e}{N_c}$
can range from $10^{-6}$--$10^{-2}$), since $\nicefrac{N_e}{N} \approx
\nicefrac{1}{V_w}$
\parencite{Hedgecock1994-gs,Wright1938-tv,Nunney1993-ef,Nunney1996-wy}, and
require multiple-merger coalescent processes  to describe their genealogies
\parencite{Eldon2006-ui}. Overall, these reproductive systems diminish the
diversity in many species, but seem unlikely to explain Lewontin's Paradox
broadly across metazoans.

Alternatively, selective processes, and in particular the indirect effects of
selection on linked neutral variation, could explain the observed narrow range
of diversity. The earliest mathematical model of hitchhiking was proffered as a
solution to Lewontin's Paradox \parencite{Maynard_Smith1974-zr}.  Since,
empirical observations have demonstrated that linked selection shapes patterns
of genome-wide diversity, as evidenced by the correlation between recombination
and diversity in a variety of species
\parencite{Aguade1989-jx,Begun1992-ey,Cutter2003-tl,Stephan1998-hh,Cai2009-by}.
Theoretic work to explain this pattern has considered diversity under a steady
influx of new beneficial mutations (recurrent hitchhiking;
\cite{Stephan1992-jc,Stephan1995-ry}), and purifying selection against new
deleterious mutations (background selection, BGS;
\cite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}).
Indeed, empirical work indicates background selection diminishes diversity
around genic regions in a variety of species
\parencite{McVicker2009-ax,Hernandez2011-gs,Charlesworth1996-px}, and now
efforts have shifted towards teasing apart the effects of positive and negative
selection on genomic diversity \parencite{Elyashiv2016-vt}.

An important class of theoretic selection models pertaining to Lewontin's
Paradox are recurrent hitchhiking models that decouple diversity from the
census population size. These models predict diversity when strongly selected
beneficial mutations regularly enter and sweep through the population, trapping
lineages and forcing them to coalesce
\parencite{Kaplan1989-sc,Gillespie2000-mh}. In general, decoupling occurs under
these hitchhiking models when the rate of coalescence due to selection is much
greater than the rate of neutral coalescence \parencite[equation
22]{Coop2012-cd}. Other selection models cannot alone decouple diversity from
population size, \emph{ceteris paribus}. For example, the reduction in
diversity predicted under background selection and polygenic fitness variation
is a proportion reduction in population size, mediated by the total
recombination map length
\parencite{Charlesworth1993-gb,Nicolaisen2012-vs,Nordborg1996-nq,Robertson1961-ho,Santiago1995-hx,Santiago1998-bs}.

@@

Recently, \textcite{Corbett-Detig2015-gt} used population genomic data to
estimate the reduction in diversity due to background selection and hitchhiking
across 40 species, and showed the impact of selection increases with two
proxies of census population size, species range and with body size. These
authors argued this is evidence that selection could explain Lewontin's
Paradox; however, in a re-analysis, \textcite{Coop2016-gx} demonstrated that
the observed scale of these reductions is insufficient to explain the
orders-of-magnitude shortfall between observed and expected levels of diversity
across species. Other recent work has found that certain life history
characteristics related to parental investment, such as propagule size, are
good predictors diversity in animals \parencite{Romiguier2014-bp,Chen2017-nf}.
Nevertheless, while these diversity correlates are important clues, they do not
propose a mechanism by which these traits act to constrain diversity within a
few orders of magnitude.

Here, I revisit Lewontin's Paradox by integrating a variety of data sets and
assessing the predicted reductions in diversity under different selection
models. Prior surveys of genetic diversity either lacked census population size
estimates, used allozyme-based measures of heterozygosity, or included fewer
species. To address these shortcomings, I first estimate census sizes by
combining predictions of population density based on body size with ranges
estimated from geographic occurrence data. Using these estimates, I quantify
the relationship between census size and previously-published genomic diversity
estimates across 172 metazoan taxa within nine phyla, which provides a sense of
the scale of the divergence between $\pi$ and $N_c$ that leads to Lewontin's
Paradox.

Past work looking at the relationship between $\pi$ and $N_c$ has largely
ignored phylogenetic non-independence across taxa
\parencite{Felsenstein1985-an}. To address this shortcoming, I account for
phylogenetic non-independence across taxa using a synthetic time-calibrated
phylogeny and phylogenetic comparative methods (PCMs). Moreover,
\textcite{Lynch2011-qv} has argued that since coalescent times are much less
than divergence times, considering phylogenetic non-independence is unnecessary
for traits like effective population size. Using PCMs, I test this conjecture
by estimating phylogenetic signal in the diversity census size relationship,
and investigating how these traits evolve along the phylogeny.

Finally, I explore whether the predicted reductions of diversity under
background selection and recurrent hitchhiking are sufficiently strong 
to resolve Lewontin's Paradox. These predicted reductions in diversity across
species are generously estimated using selection strength parameters from
\emph{Drosophila melanogaster}, a species known to be strongly affected by
linked selection. Given the effects of linked selection are mediated by
recombination map length, I investigate how recombination map lengths vary with
census population size using data from a previously-published survey
\parencite{Stapley2017-fs}. I find map lengths are typically shorter in
large--census-size species, increasing the effects of linked selection in these
species, which might further decouple diversity from census size. Still, I find
the combined impact of these selection models with available parameter
estimates falls short in explaining Lewontin's Paradox, and discuss future
avenues through which the Paradox of Variation could be fully resolved.

@@

A major impediment in quantifying Lewontin's Paradox has been estimating census
population sizes across many taxa, especially for extremely abundant,
cosmopolitan species that define the upper limit of ranges. Previous work has
surveyed the literature for census size estimates
\parencite{Nei1984-zi,Soule1976-he,Frankham1996-yb}, or used range and body
size or qualitative categories as proxies for census size
\parencite{Corbett-Detig2015-gt,Leffler2012-zj}. To quantify the relationship
between genomic estimates of diversity and census population sizes, I first
approximate census population sizes for 172 metazoan taxa (Figure
\ref{fig:figure-nc}). My approach predicts population densities from body sizes
using a previously-observed linear relationship that holds across metazoans
(Supplementary Figure \ref{suppfig:damuth};
\cite{Damuth1981-st,Damuth1987-sg}). Then, from geographic occurrence data, I
estimate range sizes. Finally, I estimate population size as the product of
these predicted densities and range estimates (see Methods:
\Newnameref{sec:methods-popsize}). Note that the relationship between
population density and body size is driven by energy budgets, and thus reflects
macroecological equilibria \parencite{Damuth1987-sg}; consequently, population
sizes for taxa like humans and their domesticated species are underestimated.
While these methods to estimate census size are crude and approximate, they can
be efficiently calculated for numerous taxa and are sufficient to estimate the
scale of Lewontin's Paradox.

@@

To determine which ecological or evolutionary processes could decouple
diversity from census population size, we first need to quantify this
relationship across a wide variety of taxa. Previous work has found there is a
significant relationship between heterozygosity and the logarithm of population
size, but these studies relied on heterozygosity measured from allozyme data
\parencite{Soule1976-he,Frankham1996-yb,Nei1984-zi}. Here, I confirm these
findings using pairwise diversity estimates from genomic sequence data and the
estimated census sizes (Figure \ref{fig:figure-1}). The pairwise diversity
estimates are from three sources: \textcite{Leffler2012-zj},
\textcite{Corbett-Detig2015-gt}, and \textcite{Romiguier2014-bp}, and are
predominantly from either synonymous or non-coding DNA (see Methods:
\Newnameref{sec:methods-diversity}). Overall, an ordinary least squares (OLS)
relationship on a log-log scale fits the data well (Figure \ref{fig:figure-1}).
The OLS slope estimate is significant and implies a 0.09 percent increase in
differences per basepair for every order of magnitude census size grows (95\%
confidence interval $[0.08, 0.12]$; see also the OLS fit per-phyla,
Supplementary Figure \ref{suppfig:figure-1-ave}).

@@

In quantifying Lewontin's Paradox, I have additionally fit some simple trait
evolution models that account for phylogenetic non-independence, investigated
whether there is a signal of phylogenetic non-independence, estimated the
continuous trait values on the phylogeny, and assessed how diversity and
population size evolve. Prior population genetic comparative studies have
lacked time-calibrated phylogenies and assumed unit branch lengths
\parencite{Whitney2010-ud}, a shortcoming that has drawn criticism
\parencite{Lynch2011-qv}. Here, I use a synthetic time-calibrated phylogeny
created from the DateLife project \parencite{OMeara2020-ds} to account for
shared phylogenetic history (see Methods: \Newnameref{sec:methods-pcm}).

@@

Using a phylogenetic mixed-effects model
\parencite{Lynch1991-tp,Hadfield2010-ql,De_Villemereuil2014-kt} implemented in
Stan \parencite{Carpenter2017-xi,Stan_Development_Team2020-ea}, I estimated the
linear relationship between diversity and population size (on a log-log scale)
accounting for phylogeny, for the 166 taxa with non-missing data and present in
the synthetic chronogram. As with the non-phylogenetic regression, this
relationship was positive and significant (95\% credible interval $0.04,
0.11$), though somewhat attenuated compared to the OLS estimates (Figure
\ref{fig:figure-2}B). Since the population size estimates are based on range
and body mass, they are essentially a composite trait; fitting phylogenetic
mixed-effects models separately on body mass and range indicates these have
significant negative and positive effects, respectively (Supplementary Figure
\ref{suppfig:figure-div-range-bodymass}).

@@

A closer visual inspection of the estimated ancestral continuous values for
diversity and population size on the phylogeny indicates the high phylogenetic
signal seems to be driven in part by chordates having low diversity and small
population sizes compared to non-chordates (Figure \ref{fig:figure-2}A). This
suggests \citeauthor{Gillespie1991-qa}'s (1991) earlier critique that the
$\pi$--$N_c$ relationship was driven by chordate-arthropod differences may be
valid. This problem resembles Felsenstein's worst-case scenario
\parencite{Felsenstein1985-an,Uyeda2018-wf}, where a singular event on a
lineage separating two clades generates a spurious association between two
traits. To further investigate whether clade-level differences dominated the
relationship between diversity and population size, I fit phylogenetic
mixed-effects models to phyla-level subsets of the data for clades with
sufficient sample sizes (see Methods: \Newnameref{sec:methods-pcm}). This
analysis shows a significant positive relationship between diversity and
population size in arthropods, and positive weak relationships in molluscs and
chordates (Supplementary Figure \ref{suppfig:div_popsize_post_phyla}). Each of
the 90\% credible intervals for slope overlap, indicating the relationship
between $\pi$ and $N_c$ is similar across these clades.

One limitation of the phylogenetic mixed-effects models employed here is that
they assume traits evolve under constant-rate Brownian motion. To test this
assumption, I performed node-height tests \parencite{Freckleton2006-ww}.
Node-height tests regress the absolute values of the standardized contrasts
between lineages against the branching time (since present) of these lineages.
Under Brownian Motion (BM), standardized contrasts are estimates of the rate of
character evolution \parencite{Felsenstein1985-an}; if a trait evolves under
constant rate BM, this relationship should be flat. For both diversity and
population size, node-height tests indicate a significant increase in the rate
of evolution towards the present (robust regression p-values 0.028 and 0.00070
respectively; Figure \ref{fig:figure-2}C).  Considering the constituents of the
population size estimate, range and body mass, separately, the rate of
evolution of range but not body mass shows a significant increase (p-value $1.9
\times 10^{-7}$) towards the present.

Interestingly, the diversity node-height test reveals two rate shifts at deeper
splits (Figure \ref{fig:figure-2}C, top left) around 570 Mya. These nodes
represent the branches between tunicates and vertebrates in chordates, and
cephalopods and pleistomollusca (bivalves and gastropods) in molluscs.  While
the cephalopod-pleistomollusca split outlier may be an artifact of having a
single cephalopod (\emph{Sepia officinalis}) in the phylogeny, the
tunicate-vertebrate split outlier is driven by the low diversity of vertebrates
and the previously-documented exceptionally high diversity of tunicates (sea
squirts; \cite{Nydam2010-kg,Small2007-mt}). This deep node representing a rate
shift in diversity could reflect a change in either effective population size
or mutation rate, and there is some evidence of both in this genus \emph{Ciona}
\parencite{Small2007-mt,Tsagkogeorga2012-lv}. Neither of these deep rate shifts
in diversity is mirrored in the population size node-height test (Figure
\ref{fig:figure-2}C, top right). Rather, it appears a trait impacting diversity
but not census size (e.g. mutation rate or offspring distributions) has
experienced a shift on the lineage separating tunicates and vertebrates. At
nearly 600 Mya, these deep nodes illustrate a counterexample to Lynch's claim
that expected coalescent times do not share phylogenetic history because they
are less than divergence times.

@@

>>>>

% TODO recheck node height and other p-values

@@

\begin{abstract}

  Under neutral theory, genetic diversity in an equilibrium population is
  expected to increase with population size. However, observed levels of
  diversity across metazoans vary only two orders of magnitude, while census
  population sizes ($N_c$) are expected to vary over several. This unexpectedly
  narrow range of diversity is a longstanding enigma in evolutionary genetics
  known as Lewontin's Paradox of Variation (1974). Since Lewontin's
  observation, it has been argued that selection constrains diversity across
  species, yet tests of this hypothesis seem to fall short of explaining the
  orders-of-magnitude reduction in diversity observed in nature. In this work,
  I revisit Lewontin's Paradox and assess whether current models of linked
  selection are capable of constraining diversity to this extent. To quantify
  the discrepancy between pairwise diversity and census population sizes across
  species, I combine genetic data from 172 metazoan taxa with estimates of
  census sizes from geographic occurrence data and population densities
  estimated from body mass.  Next, I fit the relationship between
  previously-published estimates of genomic diversity and these approximate
  census sizes to quantify Lewontin's Paradox. To account for phylogenetic
  non-independence, I use phylogenetic comparative methods to investigate the
  diversity census size relationship, estimate phylogenetic signal, and explore
  how diversity changes along the phylogeny. I consider whether the reduction
  in diversity predicted by models of recurrent hitchhiking and background
  selection could explain the observed pattern of diversity across species.
  Since the impact of linked selection is mediated by recombination map length,
  I also investigate how map lengths vary with census sizes. I find species
  with large census sizes have shorter map lengths, leading these species to
  likely experience greater reductions in diversity due to linked selection.
  Even after using high estimates of the strength of sweeps and background
  selection, I find linked selection likely cannot explain the shortfall
  between predicted and observed diversity levels across metazoan species.
  Furthermore, the predicted diversity under linked selection does not fit the
  observed diversity--census-size relationship, implying that processes other
  than background selection and recurrent hitchhiking must be limiting
  diversity.

@@

A longstanding mystery in evolutionary genetics is that the observed levels of
genetic variation across sexual species are confined to an unexpectedly narrow
range. Under neutral theory, the average number of nucleotide differences
between lineages (pairwise diversity, $\pi$) is determined by the balance of
new mutations and their loss by genetic drift
\parencite{Kimura1964-ul,Malecot1948-zv,Wright1931-fl}. In particular, the
expected diversity at neutral sites in a panmictic population of $N_c$ diploids
is expected to be $\pi \approx 4N_c \mu$, where $\mu$ is the per basepair per
generation mutation rate. Given that metazoan germline mutation rates only
differ 10-fold ($10^{-8}$--$10^{-9}$, \cite{Kondrashov2010-fi,Lynch2010-ki}),
and census sizes vary over several orders of magnitude, one would expected
under neutral theory that heterozygosity should also vary over several orders
of magnitude. However, early allozyme surveys revealed that heterozygosity
levels across a wide range of species varied just an order of magnitude
\parencite[][p.  208]{Lewontin1974-jb}; this is known as Lewontin's ``Paradox
of Variation".  With modern sequencing-based estimates of $\pi$ across taxa
ranging over only three orders of magnitude ($0.01$--$10\%$,
\cite{Leffler2012-zj}), Lewontin's paradox has persisted unresolved through the
genomics era.

From the beginning, explanations for Lewontin's Paradox have been framed in
terms of the neutralist--selectionist controversy
\parencite{Lewontin1974-jb,Kimura1984-ia,Gillespie1991-qa,Gillespie2001-mv}.
The neutralist view is that beneficial alleles are sufficiently rare and
deleterious alleles removed sufficiently quickly, that levels of genetic
diversity are shaped predominantly by genetic drift and mutation
\parencite{Kimura1984-ia}. Specifically, \emph{non-selective} processes
decouple the effective population size implied by observed levels of diversity
$\widehat{\pi}$, $\widetilde{N}_e = \nicefrac{\widehat{\pi}}{4\mu}$, from the
census size, $N_c$. By contrast, the selectionist view is that the direct and
indirect effects of linked selection suppress diversity levels across taxa,
specifically because the impact of linked selection is greater in large
populations. Undoubtedly, these opposing views represent a false dichotomy, as
population genomic studies have uncovered both complex demographic histories
that impact diversity within a species (e.g.
\cite{Zhao2013-vd,Palkopoulou2015-bg}) and evidence that selection depresses
genome-wide diversity (e.g.
\cite{Elyashiv2016-vt,Begun1992-ey,Aguade1989-jx,McVicker2009-ax}).

@@

While chance demographic events like bottlenecks and recent expansions have
long-term impacts on diversity (since mutation-drift equilibrium is reached on
the order of size of the population), characteristics of the breeding structure
such as high variance ($V_w$) or skew in reproductive success  also suppress
diversity below the levels predicted by the census size
\parencite{Wright1938-tv}. For example, in many marine animals,
females are highly fecund, and dispersing larvae face extremely low
survivorship, leading to high variance in reproductive success
\parencite{Waples2018-kb,Waples2013-wi,Hedgecock2011-ku,Hauser2008-fd}. Such
``sweepstakes" reproductive systems can lead to remarkably small ratios of
effective population size to census population size (e.g. $\nicefrac{N_e}{N_c}$
can range from $10^{-6}$--$10^{-2}$), since $\nicefrac{N_e}{N} \approx
\nicefrac{1}{V_w}$
\parencite{Hedgecock1994-gs,Wright1938-tv,Nunney1993-ef,Nunney1996-wy}, and
require multiple-merger coalescent processes to describe their genealogies
\parencite{Eldon2006-ui}. Overall, these reproductive systems diminish the
diversity in many species, but seem unlikely to explain Lewontin's Paradox
broadly across metazoans.

Alternatively, selective processes, and in particular the indirect effects of
selection on linked neutral variation, could explain the observed narrow range
of diversity. The earliest mathematical model of hitchhiking was proffered as a
explanation of Lewontin's Paradox \parencite{Maynard_Smith1974-zr}.  Since,
empirical observations have demonstrated that linked selection shapes patterns
of genome-wide diversity, as evidenced by the correlation between recombination
and diversity in a variety of species
\parencite{Aguade1989-jx,Begun1992-ey,Cutter2003-tl,Stephan1998-hh,Cai2009-by}.
Theoretic work to explain this pattern has considered diversity under a steady
influx of new beneficial mutations (recurrent hitchhiking;
\cite{Stephan1992-jc,Stephan1995-ry}), and purifying selection against new
deleterious mutations (background selection, BGS;
\cite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}).
Indeed, empirical work indicates background selection diminishes diversity
around genic regions in a variety of species
\parencite{McVicker2009-ax,Hernandez2011-gs,Charlesworth1996-px}, and now
efforts have shifted towards teasing apart the effects of positive and negative
selection on genomic diversity \parencite{Elyashiv2016-vt}.

An important class of theoretic selection models pertaining to Lewontin's
Paradox are recurrent hitchhiking models that decouple diversity from the
census population size. These models predict diversity when strongly selected
beneficial mutations regularly enter and sweep through the population, trapping
lineages and forcing them to coalesce
\parencite{Kaplan1989-sc,Gillespie2000-mh}. In general, decoupling occurs under
these hitchhiking models when the rate of coalescence due to selection is much
greater than the rate of neutral coalescence \parencite[e.g.][equation
22]{Coop2012-cd}. Other linked selection models are proportional to population
size, and thus cannot decouple diversity, all else equal. For example, models
of background selection and polygenic fitness variation predict diversity is
proportional to population size, mediated by the total recombination map length
and the deleterious mutation rate or fitness variation
\parencite{Charlesworth1993-gb,Nicolaisen2012-vs,Nordborg1996-nq,Robertson1961-ho,Santiago1995-hx,Santiago1998-bs}.

@@

Recently, \textcite{Corbett-Detig2015-gt} used population genomic data to
estimate the reduction in diversity due to background selection and hitchhiking
across 40 species, and showed the impact of selection increases with two
proxies of census population size, species range and with body size. They
argued this is evidence that selection could explain Lewontin's Paradox;
however, in a re-analysis, \textcite{Coop2016-gx} demonstrated that the
observed scale of these reductions is insufficient to explain the
orders-of-magnitude shortfall between observed and expected levels of diversity
across species. Other recent work has found that certain life history
characteristics related to parental investment, such as propagule size, are
good predictors diversity in animals \parencite{Romiguier2014-bp,Chen2017-nf}.
Nevertheless, while these diversity correlates are important clues, they do not
propose a mechanism by which these traits act to constrain diversity within a
few orders of magnitude.

Here, I revisit Lewontin's Paradox by integrating a variety of data sets and
assessing the predicted reductions in diversity under different selection
models. Prior surveys of genetic diversity either lacked census population size
estimates, used allozyme-based measures of heterozygosity, or included fewer
species. To address these shortcomings, I first estimate census sizes by
combining predictions of population density based on body size with ranges
estimated from geographic occurrence data. Using these estimates, I quantify
the relationship between census size and previously-published genomic diversity
estimates across 172 metazoan taxa within nine phyla, to provide a sense of the
scale of the divergence between $\pi$ and $N_c$ that leads to Lewontin's
Paradox.

Past work looking at the relationship between $\pi$ and $N_c$ has been unable
to account for phylogenetic non-independence across taxa
\parencite{Felsenstein1985-an}. To address this shortcoming, I use phylogenetic
comparative methods (PCMs) with a synthetic time-calibrated phylogeny to
account for shared phylogenetic history. Moreover, it is disputed whether
considering phylogenetic non-independence is necessary in population genetics,
since coalescent times are much less than divergence times
\parencite{Whitney2010-ud,Lynch2011-qv}. Using PCMs, I address this by
estimating the degree of phylogenetic signal in the diversity census size
relationship, and investigating how these traits evolve along the phylogeny.

Finally, I explore whether the predicted reductions of diversity under
background selection and recurrent hitchhiking are sufficiently strong to
resolve Lewontin's Paradox. These predicted reductions in diversity across
species are estimated using strong selection parameters from \emph{Drosophila
melanogaster}, a species known to be strongly affected by linked selection.
Given the effects of linked selection are mediated by recombination map length,
I investigate how recombination map lengths vary with census population size
using data from a previously-published survey \parencite{Stapley2017-fs}. I
find map lengths are typically shorter in large--census-size species,
increasing the effects of linked selection in these species, which might
further decouple diversity from census size. Still, I find the combined impact
of these selection models with available parameter estimates falls short in
explaining Lewontin's Paradox, and discuss future avenues through which the
Paradox of Variation could be fully resolved.

@@

A major impediment in quantifying Lewontin's Paradox has been estimating census
population sizes across many taxa, especially for extremely abundant,
cosmopolitan species that define the upper limit of ranges. Previous work has
surveyed the literature for census size estimates
\parencite{Nei1984-zi,Soule1976-he,Frankham1996-yb}, or used range or body
size or qualitative categories as proxies for census size
\parencite{Corbett-Detig2015-gt,Leffler2012-zj}. To quantify the relationship
between genomic estimates of diversity and census population sizes, I first
approximate census population sizes for 172 metazoan taxa (Figure
\ref{fig:figure-nc}). My approach predicts population densities from body sizes
using a previously-observed linear relationship that holds across metazoans
(Supplementary Figure \ref{suppfig:damuth};
\cite{Damuth1981-st,Damuth1987-sg}). Then, from geographic occurrence data, I
estimate range sizes.  Finally, I estimate population size as the product of
these predicted densities and range estimates (see Methods:
\Newnameref{sec:methods-popsize}). Note that the relationship between
population density and body size is driven by energy budgets, and thus reflects
macroecological equilibria \parencite{Damuth1987-sg}. Consequently, population
sizes are underestimated for taxa like humans and their domesticated species,
and overestimated for species with anthropogenically reduced densities or
patchy ranges. For example, the population size of \emph{Lynx lynx} is likely
around 50,000 \parencite{Iucn2020-ap} which is around two orders of magnitude
smaller than my estimate. Additionally, the range size estimates do not
consider whether an area has unsuitable habitat, and thus may be overestimates
for species with particular niches or patchy habitats. While these methods to
estimate census size are crude and approximate, they can be efficiently
calculated for numerous taxa and are sufficient to estimate the scale of
Lewontin's Paradox.

@@

To determine which ecological or evolutionary processes could decouple
diversity from census population size, we first need to quantify this
relationship across a wide variety of taxa. Previous work has found there is a
significant relationship between heterozygosity and the logarithm of population
size or range size, but these studies relied on heterozygosity measured from
allozyme data
\parencite{Soule1976-he,Frankham1996-yb,Nei1984-zi,Leffler2012-zj}. Here, I
confirm these findings using pairwise diversity estimates from genomic sequence
data and the estimated census sizes (Figure \ref{fig:figure-1}). The pairwise
diversity estimates are from three sources: \textcite{Leffler2012-zj},
\textcite{Corbett-Detig2015-gt}, and \textcite{Romiguier2014-bp}, and are
predominantly from either synonymous or non-coding DNA (see Methods:
\Newnameref{sec:methods-diversity}). Overall, an ordinary least squares (OLS)
relationship on a log-log scale fits the data well (Figure \ref{fig:figure-1}).
The OLS slope estimate is significant and implies a 0.09 percent increase in
differences per basepair for every order of magnitude census size grows (95\%
confidence interval $[0.08, 0.12]$; see also the OLS fit per-phyla,
Supplementary Figure \ref{suppfig:figure-1-ave}).

@@

One limitation of using ordinary least squares is that shared phylogenetic
history can create correlation structure in the residuals, which violates an
assumption of the regression model
\parencite{Felsenstein1985-an,Revell2010-mf}. To address this shortcoming, I
fit this relationship using a phylogenetic mixed-effects model, investigated
whether there is a signal of phylogenetic non-independence, estimated the
continuous trait values on the phylogeny, and explored how diversity and
population size evolve. Prior population genetic comparative studies have
lacked time-calibrated phylogenies and assumed unit branch lengths
\parencite{Whitney2010-ud}, a shortcoming that has drawn criticism
\parencite{Lynch2011-qv}. Here, I use a synthetic time-calibrated phylogeny
created from the DateLife project \parencite{OMeara2020-ds} to account for
shared phylogenetic history (see Methods: \Newnameref{sec:methods-pcm}).

@@

Using a phylogenetic mixed-effects model
\parencite{Lynch1991-tp,Hadfield2010-ql,De_Villemereuil2014-kt} implemented in
Stan \parencite{Carpenter2017-xi,Stan_Development_Team2020-ea}, I estimated the
linear relationship between diversity and population size (on a log-log scale)
accounting for phylogeny, for the 166 taxa without missing data and present in
the synthetic chronogram. As with the linear regression, this relationship was
positive and significant (95\% credible interval $0.04, 0.11$), though somewhat
attenuated compared to the OLS estimates (Figure \ref{fig:figure-2}B). Since
the population size estimates are based on range and body mass, they are
essentially a composite trait; fitting phylogenetic mixed-effects models
separately on body mass and range indicates these have significant negative and
positive effects, respectively (Supplementary Figure
\ref{suppfig:figure-div-range-bodymass}; see also Supplementary Figure
\ref{suppfig:div_leffler} for the relationship between diversity and the range
categories of \cite{Leffler2012-zj}).

@@

This high degree of phylogenetic signal suggests
\citeauthor{Gillespie1991-qa}'s (1991) concern that the $\pi$--$N_c$
relationship was driven by chordate-arthropod differences may be valid. A
visual inspection of the estimated ancestral continuous values for diversity
and population size on the phylogeny indicates the high phylogenetic signal
seems to be driven in part by chordates having low diversity and small
population sizes compared to non-chordates (Figure \ref{fig:figure-2}A). This
problem resembles Felsenstein's worst-case scenario
\parencite{Felsenstein1985-an,Uyeda2018-wf}, where a singular event on a
lineage separating two clades generates a spurious association between two
traits.

To investigate whether clade-level differences dominated the relationship
between diversity and population size, I fit phylogenetic mixed-effects models
to phyla-level subsets of the data for clades with sufficient sample sizes (see
Methods: \Newnameref{sec:methods-pcm}). This analysis shows a significant
positive relationship between diversity and population size in arthropods, and
positive weak relationships in molluscs and chordates (Supplementary Figure
\ref{suppfig:div_popsize_post_phyla}). Each of the 90\% credible intervals for
slope overlap, indicating the relationship between $\pi$ and $N_c$ is similar
across these clades.

Additionally, I have explored the rate of trait change through time using
node-height tests \parencite{Freckleton2006-ww}. Node-height tests regress the
absolute values of the standardized contrasts between lineages against the
branching time (since present) of these lineages. Under Brownian Motion (BM),
standardized contrasts are estimates of the rate of character evolution
\parencite{Felsenstein1985-an}; if a trait evolves under constant rate BM, this
relationship should be flat. For both diversity and population size,
node-height tests indicate a significant increase in the rate of evolution
towards the present (robust regression p-values 0.028 and 0.00070 respectively;
Figure \ref{fig:figure-2}C). Considering the constituents of the population
size estimate, range and body mass, separately, the rate of evolution of range
but not body mass shows a significant increase (p-value $1.9 \times 10^{-7}$)
towards the present.

Interestingly, the diversity node-height test reveals two rate shifts at deeper
splits (Figure \ref{fig:figure-2}C, top left) around 570 Mya. These nodes
represent the branches between tunicates and vertebrates in chordates, and
cephalopods and pleistomollusca (bivalves and gastropods) in molluscs.  While
the cephalopod-pleistomollusca split outlier may be an artifact of having a
single cephalopod (\emph{Sepia officinalis}) in the phylogeny, the
tunicate-vertebrate split outlier is driven by the low diversity of vertebrates
and the previously-documented exceptionally high diversity of tunicates (sea
squirts; \cite{Nydam2010-kg,Small2007-mt}). This deep node representing a rate
shift in diversity could reflect a change in either effective population size
or mutation rate, and there is some evidence of both in this genus \emph{Ciona}
\parencite{Small2007-mt,Tsagkogeorga2012-lv}. Neither of these deep rate shifts
in diversity is mirrored in the population size node-height test (Figure
\ref{fig:figure-2}C, top right). Rather, it appears a trait impacting diversity
but not census size (e.g. mutation rate or offspring distributions) has
experienced a shift on the lineage separating tunicates and vertebrates. At
nearly 600 Mya, these deep nodes illustrate that expected coalescent times can
share phylogenetic history, due to phylogenetic inertia in some combination of
population size, reproductive system, and mutation rates.

@@

\section{Sensitivity Analysis}
